Eye Fitting Straight Lines in the Modern Era

### 2022-01-28 Eye-Fitting-Straight-Lines-in-the-Modern-Era.Rmd
accepted Susan's edits; added citations; indicated comments for future review

# Introduction
We all use statistical graphics, but how do we know that the graphics we use are communicating properly? When creating a graphic, we must consider the design choices most effective for conveying the intended result. For instance, we \svp{may} decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
These design choices require that we understand the \svp{perceptual} and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.

@@

\svp{We need human testing of graphics in order to draw broad conclusions, develop guidelines for graphical design, and improve graphical communication.}
Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
All of these types of tests require different levels of use and manipulation of the information being presented in the chart. 
Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], \svp{testing} participants' abilities to detect a stimulus or a difference between two stimuli. \svp{Psychophysical methods have been used to test graphical perception, as in } @vanderplas2015signs, \svp{which used the method of adjustment} to estimate the magnitude of the impact of the sine illusion. 
\svp{However, there are more modern testing methods that have been developed since the heyday of psychophysics.}

\svp{One} major development in statistical graphics \svp{which led to more advanced testing methods} is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
The grammar of graphics serves as the fundamental framework for data visualization with the notion that graphics are built from the ground up by specifying exactly how to create a particular graph from a given data set.
Visual representations are constructed through the use of “tidy data” which is characterized as a data set in which each variable is in its own column, each observation is in its own row, and each value is in its own cell [@wickham2016r].
Graphics are viewed as a mapping from variables in a data set (or statistics computed from the data) to visual attributes such as the axes, colors, shapes, or facets on the canvas in which the chart is displayed.
Software, such as Hadley Wickham’s ggplot2 [@wickham2011ggplot2], aims to implement the framework of creating charts and graphics as the grammar of graphics recommends.

\svp{Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields interesting results.} 
@buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.

\svp{While these testing methods are excellent, there is one particular subset of statistical graphics testing methods which we intend to develop further in this paper: assessing graphics by fitting statistical models "by eye".}

@@

Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective; @mosteller1981eye]. 
Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \svp{using these types of manipulations}.

@finney1951subjective \svp{used graphical testing for computational purposes: to} determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring.
The author was interested in whether one iteration of calculations was sufficient in the estimation of parameters connected with dose-response relationships. 
One measure of interest is the relative potency between a test preparation of doses and standard preparation of does; relative potency is calculated as the ratio of two equally effective doses between the two preparation methods. 
In this study, twenty-one scientists were recruited via postal mail and asked to "rule two lines" in order to judge by eye the positions for a pair of parallel probit regression lines in a biological assay.
The author then computed one iterative calculation of the relative potency based on starting values as indicated by the pair of lines provided by each participant and compared these relative potency estimates to that which was estimated by the full probit technique (reaching convergence through multiple iterations).
Results indicated that one cycle of iterations for calculating the relative potency was sufficient based on the starting values provided by eye from the participants.

@@

In @ciccione2021can, participants were asked to judge trends, estimate slopes, and conduct extrapolation.
To estimate slopes, participants were asked to report the slope of the best-fitting regression line using a track-pad to adjust the tilt of a line on screen.
Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points. 
This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot. \svp{Cite deming regression XXX}

\svp{While not explicitly intended for perceptual testing,} in 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
The New York Times team utilizes Data Driven Documents (D3) \svp{XXX cite the D3 paper} that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse. 
After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
\svp{While this interactive feature is designed to get readers to confront their own intuitions about data in the news, we feel that the interactivity of this method may be useful for another purpose.}

In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing, \svp{validating the 'You Draw It' method} by replicating the study conducted by @mosteller1981eye. 
Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than an ordinary least squares regression.
In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.

@@

Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
A total of 35 individuals completed 119 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
All participants had normal or corrected to normal vision and signed an informed consent form.
The experimental tasks took approximately 15 minutes to complete.
While \svp{this study does utilize a convenience sample, as} this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
\svp{These data were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.} 
Participants completed the experiment on their own computers in an environment of their choosing. The experiment was conducted and distributed through a Shiny application \svp{XXX Cite Shiny} found [here](https://shiny.srvanderplas.com/you-draw-it/). \svp{We should probably make a rstudioconnect version that saves data to e.g. a google spreadsheet? Just in case my server goes down or something?}

@@

In the study, participants are shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line. After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task.
Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' visual. 
In order to allow for user interaction and data collection, we integrate the D3 visual into Shiny using the `r2d3` package.
\svp{While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.} 
<!-- Details of the development of the 'You Draw It' task plots will be addressed in future work. -->

@@

This experiment was conducted as part of a larger study \svp{of the perception of log and linear scales}; for simplicity, we focus on the study design and methods related to the current study. 
Each scatter-plot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - \svp{in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in}. 
During the practice session, participants are provided with instruction prompts accompanied by a .gif and a practice plot.
Instructions guide participants to start at the edge of the yellow box, to make sure the yellow region is moving along with their mouse as they draw, and that they can draw over their already drawn line.
Practice plots are then followed by four 'You Draw It' task plots associated with the current study.
The order of the task plots was randomly assigned for each individual in a completely randomized design.

@@

+ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data). 
+ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found [here](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
+ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://shiny.srvanderplas.com/you-draw-it/).

>>>>

# Introduction
We all use statistical graphics, but how do we know that the graphics we use are communicating properly? When creating a graphic, we must consider the design choices most effective for conveying the intended result. For instance, we may decide to highlight the relationship between two variables in a scatter-plot by including a trend line, or adding color to highlight clustering [@vanderplas2017clusters].
These design choices require that we understand the perceptual and visual biases that come into play when creating graphics, and as graphics are evaluated visually, we must use human testing to ground our understanding in empiricism.

@@

We need human testing of graphics in order to draw broad conclusions, develop guidelines for graphical design, and improve graphical communication.
Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations.
All of these types of tests require different levels of use and manipulation of the information being presented in the chart. 
Early researchers studied graphs from a psychological perspective [@spence1990visual; @lewandowsky1989perception], testing participants' abilities to detect a stimulus or a difference between two stimuli. Psychophysical methods have been used to test graphical perception, as in @vanderplas2015signs, which used the method of adjustment to estimate the magnitude of the impact of the sine illusion. 
However, there are more modern testing methods that have been developed since the heyday of psychophysics.

One major development in statistical graphics which led to more advanced testing methods is Wilkinson’s Grammar of Graphics [@wilkinson2013grammar].
The grammar of graphics serves as the fundamental framework for data visualization with the notion that graphics are built from the ground up by specifying exactly how to create a particular graph from a given data set.
Visual representations are constructed through the use of “tidy data” which is characterized as a data set in which each variable is in its own column, each observation is in its own row, and each value is in its own cell [@wickham2016r].
Graphics are viewed as a mapping from variables in a data set (or statistics computed from the data) to visual attributes such as the axes, colors, shapes, or facets on the canvas in which the chart is displayed.
Software, such as Hadley Wickham’s ggplot2 [@wickham2011ggplot2], aims to implement the framework of creating charts and graphics as the grammar of graphics recommends.

Combining the grammar of graphics with another tool for statistical graphics testing, the statistical lineup, yields interesting results.
@buja2009statistical introduced the lineup protocol to provide a framework for inferential testing.
Through experimentation, methods such as the lineup protocol allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation [@vanderplas2017clusters; @vanderplas2015spatial; @hofmann2012graphical].
The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.

While these testing methods are excellent, there is one particular subset of statistical graphics testing methods which we intend to develop further in this paper: assessing graphics by fitting statistical models "by eye".

@@

Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective; @mosteller1981eye]. 
Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations \svp{using these types of manipulations.} \ear{XXX this study was conducted online, but asked users to use the arrows on their keyboard to adjust the slope or select whether the shown slope was positive or negative}.

@finney1951subjective used graphical testing for computational purposes: to determine the effect of stopping iterative maximum likelihood calculations after one iteration. 
Many techniques in statistical analysis are performed with the aid of iterative calculations such as Newton's method or Fisher's scoring.
The author was interested in whether one iteration of calculations was sufficient in the estimation of parameters connected with dose-response relationships. 
One measure of interest is the relative potency between a test preparation of doses and standard preparation of does; relative potency is calculated as the ratio of two equally effective doses between the two preparation methods. 
In this study, twenty-one scientists were recruited via postal mail and asked to "rule two lines" in order to judge by eye the positions for a pair of parallel probit regression lines in a biological assay.
The author then computed one iterative calculation of the relative potency based on starting values as indicated by the pair of lines provided by each participant and compared these relative potency estimates to that which was estimated by the full probit technique (reaching convergence through multiple iterations).
Results indicated that one cycle of iterations for calculating the relative potency was sufficient based on the starting values provided by eye from the participants.

@@

In @ciccione2021can, participants were asked to judge trends, estimate slopes, and conduct extrapolation.
To estimate slopes, participants were asked to report the slope of the best-fitting regression line using a track-pad to adjust the tilt of a line on screen.
Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increase with noise and with number of points. 
This supports the results found in @mosteller1981eye and suggest that participants might use Deming regression when fitting a line to a noisy scatter-plot [@deming1943statistical; @linnet1998performance; @martin2000general] \ear{xxx added citation}.

While not explicitly intended for perceptual testing, in 2015, the New York Times introduced an interactive feature, called You Draw It [@aisch_cox_quealy_2015; @buchanan_park_pearce_2017; @katz_2017].
Readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
The New York Times team utilizes Data Driven Documents (D3) [@bostock2011d3] \ear{xxx added citation} that allows readers to predict these metrics through the use of drawing a line on their computer screen with their computer mouse. 
After the reader has completed drawing the line, the actual observed values are revealed and the reader may check their estimated knowledge against the actual reported data.
While this interactive feature is designed to get readers to confront their own intuitions about data in the news, we feel that the interactivity of this method may be useful for another purpose.

In this paper, we establish 'You Draw It', adapted from the New York Times feature, as a tool for graphical testing, validating the 'You Draw It' method by replicating the study conducted by @mosteller1981eye. 
Based on previous research surrounding "ensemble perception," we hypothesize that visual regression tends to mimic principle component regression rather than an ordinary least squares regression.
In order to assess this hypothesis, we introduce a method for statistically modeling the participant drawn lines using generalized additive mixed models.

@@

Participants were recruited through through Twitter, Reddit, and direct email in May 2021.
A total of 35 individuals completed 119 unique 'You Draw It' task plots; all completed you draw it task plots were included in the analysis. 
All participants had normal or corrected to normal vision and signed an informed consent form.
The experimental tasks took approximately 15 minutes to complete.
While this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
These data were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.
Participants completed the experiment on their own computers in an environment of their choosing. The experiment was conducted and distributed through a Shiny application [@shinyPkg] \ear{xxx added citation} found [here](https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/). \svp{We should probably make a rstudioconnect version that saves data to e.g. a google spreadsheet? Just in case my server goes down or something?} \ear{XXX currently linked to my shinyapps.io page but will get this updated to a new repository.}

@@

In the study, participants are shown an interactive scatter-plot \pcref{fig:ydi-stimuli} along with the prompt, “Use your mouse to fill in the trend in the yellow box region.”
The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line. After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task.
Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' visual. 
In order to allow for user interaction and data collection, we integrate the D3 visual into Shiny using the `r2d3` package.
While the interface is highly customized to this particular task, we hope to generalize the code and provide a Shiny widget in an R package soon.

@@

This experiment was conducted as part of a larger study of the perception of log and linear scales; for simplicity, we focus on the study design and methods related to the current study. Each scatter-plot was the graphical representation of a data set that was generated randomly and independently for each participant at the start of the experiment.
Participants in the study are shown two 'You Draw It' practice plots in order to train participants in the skills associated with executing the task - in particular, the responsiveness of the applet requires that participants draw a line at a certain speed, ensuring that all of the evenly spaced points along the hand-drawn line are filled in. 
During the practice session, participants are provided with instruction prompts accompanied by a .gif and a practice plot.
Instructions guide participants to start at the edge of the yellow box, to make sure the yellow region is moving along with their mouse as they draw, and that they can draw over their already drawn line.
Practice plots are then followed by four 'You Draw It' task plots associated with the current study.
The order of the task plots was randomly assigned for each individual in a completely randomized design.

@@

+ **Participant Data:** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub [here](https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/tree/main/data). 
+ **Data Analysis Code:** The code used to replicate the analysis in this paper can be found [here](https://earobinson95.github.io/Eye-Fitting-Straight-Lines-in-the-Modern-Era/analysis/you-draw-it-eyefitting-analysis.html).
+ **Study Applet:** The shiny app used to conduct the study can be accessed [here](https://emily-robinson.shinyapps.io/you-draw-it-pilot-app/).
+ **\ear{RShiny Applet Code:}** \ear{The code used to create the RShiny Applet for data collection can be found} [\ear{here}](https://github.com/earobinson95/you-draw-it-validation-applet).
